\fontsize{18}{22}\selectfont A comprehensive study of the phylodynamics of SARS-CoV-2 in Europe

### 2021-03-05 reports/mt-thesis/methods.tex
wip methods almost finished

\textbf{SARS-CoV 2 genome data set}

We create a dataset of 360 genetic sequences from December 2019 to March 8 2020, obtained from publicly available data on GISAID \cite{gisaid} (accessed on November 2020). We follow the Nextstrain workflow for the curation of the dataset \cite{nextstrain}. Sequences with incomplete collection date, less than 27.000 bases in length or with more than 3.000 unknown bases are omitted. Also, sequences from known clusters of transmission or from the same patient are excluded. The resulting dataset of x \todo{look number of seqs} sequences is aligned with MAFFT. The begining and the end of the alignment are masked respectively by 100 and 50 sequences as well as sites  13402, 24389 and 24390, identified by Nextstrain as prone to sequencing errors.

@@

\textbf{The multitype birth death model}

To study the early dynamics of SARS-CoV-2 in the European countries, we use a simplified version of the multitype birth-death model described in \cite{Kuhnert2016}, following the analysis in \cite{Nadeau2020}. Birth death models are compartmental population models with high flexibility that describe the process of epidemiological transmission. The stochastic formulation of these models are used in phylodynamic analyses \cite{Stadler2012}. In the multitype version, we consider a structured population in types or subpopulations with characteristic within-subpopulation dynamics and migrations between them. In our case, the subpopulations are the different locations of the samples.

@@

\textbf{Epidemic trajectories and structured trees}

We will refer to the full sequence of transmissions, recoveries/deaths, migrations and sampling events as an epidemic trajectory. One set of sequences, and therefore one phylogenetic tree, is the product of one epidemic trajectory. Moreover, the tree represents a fraction of the events in the epidemic trajectory, those that involve the sampled individuals. In epidemiology, we aim to learn about the true epidemic trajectory since we usually have incomplete information caused by unreported cases or unknown transmission chains.

@@

\textbf{Bayesian Inference with BDMM-Prime}

We want to infer the epidemic trajectories under the multitype birth-death model fit to a set of samples collected throughout SARS-CoV-2 early epidemics in Europe. In order to do this, we have to estimate the joint posterior distribution of the structured tree $\mathcal{T}_c$, the epidemic trajectory $\mathcal{E}$, the substitution model parameters $\theta$ and the multi-type birth death parameters $\eta$. This posterior probability can be expressed with Bayes' formula as:

@@

@@

While it is also possible to include $\mathcal{T}_c$ and $\mathcal{E}$ directly into the Bayes' rule as implemented in EpiInf package\cite{Vaughan2019}, the factorization of the posterior in these three terms allow us to use the standard birth-death-sampling tree prior implementations in BDMM \cite{Kuhnert2016} \cite{Scire2020} to compute $P(\mathcal{T}, \theta, \eta | A)$. This will speedup the analysis with almost no overhead compared to the standard BDMM inference and can be used in pre-existing multitype analyses withoud additional MCMC.\\

\textbf{Stochastic mapping of ancestral locations and epidemic trajectories}

To sample from $P(\mathcal{T}_c|\mathcal{T}, \eta)$, BDMM-Prime implements a stochastic mapping algorithm based on the work by \cite{Hohna2019}. A set of differential equations is numerically integrated over $\mathcal{T}$ to obtain the marginal probability of a location value at any point along each branch. From this probabilities, we can derive time-dependent rates that define the changes of the location along the tree according to a continuous-time Markov process. Then, we can simulate forward in time from the root of the tree the location trajectories down the tree edges. For a detailed derivation of the stochastic mapping of the algorithm refer to \cite{VaughanWIP}.\\

@@

\textbf{Incorporating air travel data and geographic distances}

In order to inform our migration rates with external information, and to avoid having to estimate a large number of migration rates we implement a generalized linear model (GLM) for the migration matrix based on \cite{Lemey2014} and \cite{Nicola2019}. This GLM model parametrizes the migration rates as a log linear function of a number of potential predictors, Equation \ref{eq:glm}. For each predictor $x_i$, the GLM parameterization includes a coefficient $\beta_i$, which quantifies the effect size of the predictor, and a binary indicator variable $\delta_i$, that allows the predictor to be included or excluded from the model. The parameter $c$ represents the overal magnitude of migration. This means that if every indicator is 0, every migration rate will be equal to this parameter. In the MCMC, we estimate the effect size of each of the predictors, as well as their inclusion probability $\mathbb{E}[\delta_i]$.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{figures/flight_data.png}
    \caption{\todo[inline]{todo}}
    \label{fig:epitrajs}
\end{figure}

\textbf{Model specifications and priors}

Since we focus on the early epidemic outbreak, we expect un unimpeded spread of the virus that can be described by the exponential growth of the infected population and no significant decrease in the number of susceptibles over time \cite{Boskova2014}. Thus, we assume a constant basic reproductive number $R_0$ for each of the European countries before the lockdown in the Lombardy region in March 8. However, in China the first wave of the epidedmic had almost concluded in March so we can not make the same assumption. We fixed the reproductive number in China and include three different periods, changing on January 23 and February 10 based on \cite{Pan2020} \cite{Wu2020} \cite{Xiao2020}. Before January 23, we use the estimation in \cite{Park2020} \cite{Billah2020} to fix the basic reproductive number $R_0$ in China to $2.9$ . From January 23 to February 10 and from February 11 to March 8, we calculate the average effective reproductive number based on the estimations in \cite{covidre} \cite{Huisman2020} and obtain the $R_e$ in China to be $1.1$ and $0.43$ respectively.

@@

\bottomrule
\end{tabular}
\caption{Caption. Source of case data. \todo[inline]{TODO. Vertical alignment}}
\label{table:priors}
\end{table}

\textbf{Data availability}
The SARS-CoV-2 genetic data can be downloaded from www.gisaid.org. Supplementary Table \ref{stable:seqs} lists accession numbers for the genetic sequences used in this study. Reported SARS-CoV-2 case counts are obtained from the country official agencies indicated in Table \ref{table:cases}. Flight data can be obtained from EUROSTATS https://ec.europa.eu/eurostat/.

\textbf{Code availability and analyses reproducibility}
All the code is available at ...
To ensure the reproducibility of the analysis, we have implemented the whole workflow with Snakemake \cite{snakemake}. The workflow is a modified version of Nextrain workflow \cite{nextstrain} with the additional rules for BEAST 2 analysis and results processing.

>>>>

\textbf{SARS-CoV 2 genome data set.} 
We create a dataset of 360 genetic sequences from December 2019 to March 8 2020, obtained from publicly available data on GISAID \cite{Shu2017} (accessed on November 2020). We follow the Nextstrain workflow for the curation of the dataset \cite{nextstrain}. Sequences with incomplete collection date, less than 27.000 bases in length or with more than 3.000 unknown bases are omitted. Also, sequences from known clusters of transmission or from the same patient are excluded. The resulting worldwide sequence dataset is aligned with MAFFT. The begining and the end of the alignment are masked respectively by 100 and 50 sequences as well as sites  13402, 24389 and 24390, identified by Nextstrain as prone to sequencing errors.

@@

\textbf{The multitype birth death model.} 
To study the early dynamics of SARS-CoV-2 in the European countries, we use a simplified version of the multitype birth-death model described in \cite{Kuhnert2016}, following the analysis in \cite{Nadeau2020}. Birth death models are compartmental population models with high flexibility that describe the process of epidemiological transmission. The stochastic formulation of these models are used in phylodynamic analyses \cite{Stadler2012}. In the multitype version, we consider a structured population in types or subpopulations with characteristic within-subpopulation dynamics and migrations between them. In our case, the subpopulations are the different locations of the samples.

@@

\textbf{Epidemic trajectories and structured trees.} 
We will refer to the full sequence of transmissions, recoveries/deaths, migrations and sampling events as an epidemic trajectory. One set of sequences, and therefore one phylogenetic tree, is the product of one epidemic trajectory. Moreover, the tree represents a fraction of the events in the epidemic trajectory, those that involve the sampled individuals. In epidemiology, we aim to learn about the true epidemic trajectory since we usually have incomplete information caused by unreported cases or unknown transmission chains.

@@

\textbf{Bayesian Inference with BDMM-Prime.} 
We want to infer the epidemic trajectories under the multitype birth-death model fit to a set of samples collected throughout SARS-CoV-2 early epidemics in Europe. In order to do this, we have to estimate the joint posterior distribution of the structured tree $\mathcal{T}_c$, the epidemic trajectory $\mathcal{E}$, the substitution model parameters $\theta$ and the multi-type birth death parameters $\eta$. This posterior probability can be expressed with Bayes' formula as:

@@

$P(\mathcal{T}_c|\mathcal{T}, \eta)$, $P(\mathcal{E}|\mathcal{T}_c, \eta)$ and $P(\mathcal{T}, \theta, \eta | A)$ are the posterior probabilities of the structured tree, the epidemic trajectory and the joint posterior of the phylogenetic tree (without ancestral locations) and the model parameters. The tree likelihood $P(A | \mathcal{T},\theta )$ is the probability of the sequence alignment and can be efficiently evaluated using Felsenstein’s pruning algorithm \cite{Felsenstein1981}. The tree prior, $P(\mathcal{T} | \eta)$ also called phylodynamic likelihood is derived from the multitype birth-death model. $P(\theta)$ and $P(\eta)$ represent our prior belief in the distribution of the population and substitution model parameters.

@@

While it is also possible to include $\mathcal{T}_c$ and $\mathcal{E}$ directly into the Bayes' rule as implemented in EpiInf package \cite{Vaughan2019}, the factorization of the posterior in these three terms allow us to use the standard birth-death-sampling tree prior implementations in BDMM to compute $P(\mathcal{T}, \theta, \eta | A)$ \cite{Kuhnert2016} \cite{Scire2020}. This will speedup the analysis with almost no overhead compared to the standard BDMM inference and can be used in pre-existing multitype analyses withoud additional MCMC.\\

\textbf{Stochastic mapping of ancestral locations and epidemic trajectories}
To sample from $P(\mathcal{T}_c|\mathcal{T}, \eta)$, BDMM-Prime implements a stochastic mapping algorithm based on the work by \cite{Hohna2019}. A set of differential equations is numerically integrated over $\mathcal{T}$ to obtain the marginal probability of a location value at any point along each branch. From this probabilities, we can derive time-dependent rates that define the changes of the location along the tree according to a continuous-time Markov process. Then, we can simulate forward in time from the root of the tree the location trajectories down the tree edges. For a detailed derivation of the stochastic mapping of the algorithm refer to \cite{VaughanWIP}.\\

@@

\textbf{Incorporating air travel data and geographic distances.}
 In order to inform our migration rates with external information, and to avoid having to estimate a large number of migration rates we implement a generalized linear model (GLM) for the migration matrix based on \cite{Lemey2014} and \cite{Nicola2019}. This GLM model parametrizes the migration rates as a log linear function of a number of potential predictors, Equation \ref{eq:glm}. For each predictor $x_i$, the GLM parameterization includes a coefficient $\beta_i$, which quantifies the effect size of the predictor, and a binary indicator variable $\delta_i$, that allows the predictor to be included or excluded from the model. The parameter $c$ represents the overal magnitude of migration. This means that if every indicator is 0, every migration rate will be equal to this parameter. In the MCMC, we estimate the effect size of each of the predictors, as well as their inclusion probability $\mathbb{E}[\delta_i]$.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{figures/flight_data.png}
    \caption{\todo[inline]{improve figure}}
    \label{fig:epitrajs}
\end{figure}

\textbf{Model specifications and priors.} 
Since we focus on the early epidemic outbreak, we expect un unimpeded spread of the virus that can be described by the exponential growth of the infected population and no significant decrease in the number of susceptibles over time \cite{Boskova2014}. Thus, we assume a constant basic reproductive number $R_0$ for each of the European countries before the lockdown in the Lombardy region in March 8. However, in China the first wave of the epidedmic had almost concluded in March so we can not make the same assumption. We fixed the reproductive number in China and include three different periods, changing on January 23 and February 10 based on \cite{Pan2020} \cite{Wu2020} \cite{Xiao2020}. Before January 23, we use the estimation in \cite{Park2020} \cite{Billah2020} to fix the basic reproductive number $R_0$ in China to $2.9$ . From January 23 to February 10 and from February 11 to March 8, we calculate the average effective reproductive number based on the estimations in \cite{covidre} \cite{Huisman2020} and obtain the $R_e$ in China to be $1.1$ and $0.43$ respectively.

@@

\bottomrule
\end{tabular}
\caption{Caption \todo[inline]{finish table and caption}}
\label{table:priors}
\end{table}

\textbf{Data availability.}
The SARS-CoV-2 genetic data can be downloaded from www.gisaid.org \cite{Shu2017}. Supplementary Table \ref{stable:seqs} \todo{todo} lists accession numbers for the genetic sequences used in this study. Reported SARS-CoV-2 case counts are obtained from the country official agencies indicated in Table \ref{table:cases} \todo{todo}. Flight data can be obtained from EUROSTATS https://ec.europa.eu/eurostat/.

\textbf{Code availability and analyses reproducibility.}
All the code is available at ...
To ensure the reproducibility of the analysis, we have implemented the whole workflow with Snakemake \cite{snakemake}. The workflow is a modified version of Nextrain workflow \cite{nextstrain} with the additional rules for BEAST 2 analysis and results processing.
